Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms

### 2013-12-03 CGP Journal/main.tex
Don't need white space things anymore
Minor text fixes found during full readthrough.

From Figure~\ref{fig:prob} it is clear that nodes near the beginning of the genome
are almost certainly going to be active, while nodes in the last three quarters
of the genome are almost never going to be active.
\end{comment}
%NEW
Substituting $N=2000$, $A=2$, $I=6$, $O=6$ into Equation~\ref{eq:prob}
predicts only 160 active nodes, with 80\% of the first 100 nodes expected to be
active while just 4\% of the remaining 1900 nodes are expected to be active.
%/NEW

@@

The probability of connecting $X$ to $Y$ is also influenced by how many nodes transitively depend on $X$ and $Y$.
For example, assume $X$ has more nodes transitively dependent on it than $Y$.
In \emph{Reorder}, $X$ is again more likely to be placed before $Y$, this time because the number
of possible nodes preceding it is reduced.  \emph{DAG} is biased in the reverse, such that $X$ will
have a higher chance of connecting to $Y$ that $Y$ to $X$.  This is because $X$ can connect to fewer
nodes, so each mutation has a higher chance of choosing $Y$.

@@

The first problem, 3-bit parity, represents the most
common application for
CGP~\cite{yu:2001:neutrality,miller:2006:redundancy,walker:2008:cgpmodules}.
We include this problem purely for backward comparison, as we agree that in general
it is too simple of a problem to merit its own conclusions~\cite{white:2013:bgpb}.
Yet as part of the group it may help yield understanding about how each variant
performs on simple problems.

@@

On many of the problems, many of the configurations have little to no statistical
difference from the control in their evaluations required to reach the global optimum.
On Parity and Encode it is probably safe to conclude that \emph{Single} is worse than
the control.  As these are the easiest problems tested, that difference may have
little real world meaning.  \emph{DAG} is almost certainly worse than the control on
Decode.  Its results are so much worse that we can also likely conclude that \emph{DAG}
is worse than all non \emph{DAG} configuration on that problem, even though no statistical
tests are presented here.  The control on multiply did by far the worst of all configurations,
with only \emph{Accumulate Normal} likely failing statistical significance.

@@

As a final check that our results do not somehow unfairly handicap traditional CGP,
we compared our median evaluations to success with previous publications.  The results
reported for traditional CGP in~\cite{walker:2008:cgpmodules}
for 3-bit parity required 6 times as many evaluations as \emph{Skip-Normal}.
On the 3-bit Multiplier problem using a slightly different operator set,
the same source had traditional CGP requiring 4 million evaluations to success,
as opposed to our \emph{Skip-Normal} requiring 600 thousand, also a factor of 6 improvement.
%NEW
To determine the effect the difference in function set had on results, we ran
\emph{Skip-Normal} and \emph{Skip-Reorder} using their functions but our parameters.
The former improved to 453,684 while the latter increased to 303,657, maintaining
a statistically significant difference.  These
results make sense as the operator set in~\cite{walker:2008:cgpmodules} is able
to more compactly represent the solution than our set.  This puts less strain
on \emph{Normal} to deal with complex solutions, while \emph{Reorder}'s extremely
large genome size is likely made unnecessary.  Note that as we did not tune
using their operator set, our results should be considered a lower bound on quality for
using \emph{Skip-Normal} and \emph{Skip-Reorder} to solve 3-bit Multiplier
with their operators.
%/NEW

@@

%The above paragraphs were replaced by the below paragraph.
In Section~\ref{sec:ordering} we used Equation~\ref{eq:prob} to suggest \emph{Normal}
as a strong bias on the location of active nodes regardless of fitness landscape, and that
\emph{Reorder} and \emph{DAG} would likely have less or no bias.
%NEW
The example inputs used in that section match those of \emph{Normal}'s tuned
configuration on 3-bit multiply.  The predicted number of active nodes and their
location in the genome match almost exactly
the behavior displayed in all three \emph{Normal} plots in Table~\ref{tab:never_active}.
%/NEW

@@

Each diagram describes node behavior as falling into eight distinct categories
split between active and inactive nodes.  Nodes are put into the first category
they satisfy.
\\
\\
\begin{enumerate}
\item ``Constant'' - Independent of problem input, constant nodes produce a fixed output.
\item ``Used'' - Nodes still active after applying \Call{Simplify}{}.
\item ``Useful'' - Nodes with identical semantics to a ``Used'' node.
\item ``Intron'' - Any node with semantics identical to a node that was
                   active before applying \Call{Simplify}{}, but was inactive afterward.
\item ``Explore'' - All remaining nodes.
\end{enumerate}
Used nodes are by definition only in the active portion of the genome.  Similarly
Explore nodes can only be in the inactive portion of the genome.   Note that
all never active nodes are ignored by these plots, as their behavior is meaningless.

>>>>

From Figure~\ref{fig:prob} it is clear that nodes near the beginning of the genome
are almost certainly going to be active, while nodes in the last three quarters
of the genome are almost never going to be active.
\end{comment}
%NEW
Substituting $N=2000$, $A=2$, $I=6$, $O=6$ into Equation~\ref{eq:prob} (chosen to match
experimental values used in Section~\ref{sec:analysis})
predicts only 160 active nodes, or 8\% of the total genome.  Positionally, 80\% of the first 100 nodes are expected to be
active while just 4\% of the remaining 1900 nodes are expected to be active.
%/NEW

@@

The probability of connecting $X$ to $Y$ is also influenced by how many nodes transitively depend on $X$ and $Y$.
For example, assume $X$ has more nodes transitively dependent on it than $Y$.
In \emph{Reorder}, $X$ is again more likely to be placed before $Y$, this time because the number
of possible nodes preceding it is reduced.  \emph{DAG} is biased in the reverse, such that $X$ will
have a higher chance of connecting to $Y$ than $Y$ to $X$.  This is because $X$ can connect to fewer
nodes, so each mutation has a higher chance of choosing $Y$.

@@

The first problem, 3-bit Parity, represents the most
common application for
CGP~\cite{yu:2001:neutrality,miller:2006:redundancy,walker:2008:cgpmodules}.
We include this problem purely for backward comparison, as we agree that in general
it is too simple of a problem to merit its own conclusions~\cite{white:2013:bgpb}.
Yet as part of the group it may help yield understanding about how each variant
performs on simple problems.

@@

On many of the problems, many of the configurations have little to no statistical
difference from the control in their evaluations required to reach the global optimum.
On Parity and Encode it is probably safe to conclude that \emph{Single} is worse than
the control.  As these are the easiest problems tested, that difference may have
little real world meaning.  \emph{DAG} is almost certainly worse than the control on
Decode.  Its results are so much worse that we can also likely conclude that \emph{DAG}
is worse than all non \emph{DAG} configuration on that problem, even though no statistical
tests are presented here.  The control on Multiply did by far the worst of all configurations,
with only \emph{Accumulate Normal} likely failing statistical significance.

@@

As a final check that our results do not somehow unfairly handicap traditional CGP,
we compared our median evaluations to success with previous publications.  The results
reported for traditional CGP in~\cite{walker:2008:cgpmodules}
for 3-bit Parity required 6 times as many evaluations as \emph{Skip-Normal}.
On the 3-bit Multiplier problem using a slightly different operator set,
the same source had traditional CGP requiring 4 million evaluations to success,
as opposed to our \emph{Skip-Normal} requiring 600 thousand, also a factor of 6 improvement.
%NEW
To determine the effect the difference in function set had on results, we ran
\emph{Skip-Normal} and \emph{Skip-Reorder} using their functions but our parameters.
The former improved to 453,684 while the latter increased to 303,657, maintaining
a statistically significant difference.  These
results make sense as the operator set in~\cite{walker:2008:cgpmodules} is able
to more compactly represent the solution than our set.  This puts less strain
on \emph{Normal} to deal with complex solutions, while \emph{Reorder}'s extremely
large genome size is likely made unnecessary.  Note that as we did not tune
using their operator set, our results should be considered a lower bound on quality for
using \emph{Skip-Normal} and \emph{Skip-Reorder} to solve 3-bit Multiplier
with their operators.
%/NEW

@@

%The above paragraphs were replaced by the below paragraph.
In Section~\ref{sec:ordering} we used Equation~\ref{eq:prob} to suggest \emph{Normal}
has a strong bias on the location of active nodes regardless of fitness landscape, and that
\emph{Reorder} and \emph{DAG} would likely have less or no bias.
%NEW
The example inputs used in that section match those of \emph{Normal}'s tuned
configuration on 3-bit Multiply.  The predicted number of active nodes and their
location in the genome match almost exactly
the behavior displayed in all three \emph{Normal} plots in Table~\ref{tab:never_active}.
%/NEW

@@

Each diagram describes node behavior as falling into eight distinct categories
split between active and inactive nodes.  Nodes are put into the first category
they satisfy.
\begin{enumerate}
\item ``Constant'' - Independent of problem input, constant nodes produce a fixed output.
\item ``Used'' - Nodes still active after applying \Call{Simplify}{}.
\item ``Useful'' - Nodes with identical semantics to a ``Used'' node.
\item ``Intron'' - Any node with semantics identical to a node that was
                   active before applying \Call{Simplify}{}, but was inactive afterward.
\item ``Explore'' - All remaining nodes.
\end{enumerate}
Used nodes are by definition only in the active portion of the genome.  Similarly
Explore nodes can only be in the inactive portion of the genome.   Note that
all never active nodes are ignored by these plots, as their behavior is meaningless.
